A Review of Lossless Data Compression Algorithms

### 2018-12-10 paper.tex
Update the paper and presentation.

\title{A Review of Lossless Data Compression Methods}
\author{Jack Stanek}

@@

This result places a lower bound on the \emph{efficiency} of prefix
codes in general, and suggests the sorts of data which are less
compressable, and even uncompressable (since such an input must
exist!) Since entropy of some channel $X$ with an alphabet $C$ is
defined to be $H(X) = -\sum_{c \in C} P(c)\log_2{c}$, we can
illustrate the implications of this result with a simple
example. Suppose we are attempting to compress a particular channel
$X$ which is text document in which the frequencies of every letter of
the Latin alphabet are distributed equally. (It is extremely unlikely
that a text in any human language would have this property; it is used
here to illustrate a pathological edge case in general.) Then the
average word length must be greater than
$26 * -\frac{1}{26} \log_2{\frac{1}{26}} \approx 4.7$ bits. If we were
to construct a naïve code to encode the Latin alphabet, for instance
$A=1, B=2,...$, we would require 5 bits for each letter. Shannon's
Noiseless Coding Theorem says this is more or less as good as we can
do, so for this case it is very difficult to actually compress our
input. Consider another document in which case one letter, let's say
$E$, occurs significantly more often than the others, for instance
$\frac{27}{52}$ of the time, relative to $\frac{1}{52}$ for every
other letter. Then by Shannon's Noiseless Coding Theorem, it is
possible to construct a prefix code with an average word length
greater than
$25 * -\frac{1}{52} \log_2{\frac{1}{52}} - \frac{27}{52}
\log_2{\frac{27}{25}} \approx 3.2$, meaning that we can compress such
this document approximately $33\%$ more efficiently than the uniform
one. This situation is somewhat contrived, but it serves to illustrate
the fundamental limits of prefix codes in general.

@@

To illustrate: suppose we are to encode the string ``oodles of
noodles'' using LZ77, with a window of length 8. The string is 16
characters long, requiring 16 bytes to store. The current character is
set to ``o'', the first character of the string. However, since there
are two ``o'' characters in a row, the algorithm can ``overshoot'' the
end of the window in order to efficiently implement run-length
encoding (that is, recording a character and the number of copies
folloiwng it.) Thus, \texttt{(2, 0, 'o')} is emitted, and the current
character is incremented by two. The algorithm then emits
single-character prefix 3-tuples for \texttt{d, l, e, s, ' ',} (the
space character), \texttt{o}, and \texttt{f}. At

@@

Since knowing the structure of a given input can allow the
construction of a more efficient data compression scheme, it is
worthwhile to study some compression scheme which is specifically
designed to store a particular type of data. In this secion I will
describe the structure and function of the Free Lossless Audio Codec,
known more commonly by its acronym FLAC, a popular format for
high-fidelity audio storage. \cite{flac}

>>>>

\title{A Review of Lossless Data Compression Methods}
\author{Jack Stanek}
\date{December 10, 2018}

@@

This result places a lower bound on the \emph{efficiency} of prefix
codes in general, and suggests the sorts of data which are less
compressable, and even uncompressable (since such an input must
exist!) Since entropy of some channel $X$ with an alphabet $C$ is
defined to be $H(X) = -\sum_{c \in C} P(c)\log_2{c}$, we can
illustrate the implications of this result with a simple
example. Suppose we are attempting to compress some arbitrary channel,
$X$, which is text document in which the frequencies of every letter
of the Latin alphabet are distributed equally. (It is extremely
unlikely that a text in any human language would have this property;
it is used here to illustrate a pathological edge case in general.)
Then the average word length must be greater than
$26 * -\frac{1}{26} \log_2{\frac{1}{26}} \approx 4.7$ bits. If we were
to construct a naïve code to encode the Latin alphabet, for instance
$A=1, B=2,...$, we would require 5 bits for each letter. Shannon's
Noiseless Coding Theorem says this is more or less as good as we can
do, so for this case it is very difficult to actually compress our
input. Consider another document in which case one letter, let's say
$E$, occurs significantly more often than the others, for instance
$\frac{27}{52}$ of the time, relative to $\frac{1}{52}$ for every
other letter. Then by Shannon's Noiseless Coding Theorem, it is
possible to construct a prefix code with an average word length
greater than
$25 * -\frac{1}{52} \log_2{\frac{1}{52}} - \frac{27}{52}
\log_2{\frac{27}{25}} \approx 3.2$, meaning that we can compress such
this document approximately $33\%$ more efficiently than the uniform
one. This situation is somewhat contrived, but it serves to illustrate
the fundamental limits of prefix codes in general.

@@

To illustrate: suppose we want to encode the laughing string
``hahahahaha'' using the LZ77 algorithm with an 8-character window. We
would begin by encoding the first two (unique) characters as
\texttt{(0, 0, 'h')} and \texttt{(0, 0, 'a')}, respectively. Then, the
rest of the word can be encoded as \texttt{(2, 8, '\textbackslash
  0')}, where \texttt{'\textbackslash 0'} represents the ASCII null
character. The resultant output can be stored in 9 bytes, compared to
the 11 required by the original input string. This example also
illustrates how the algorithm can use a powerful implementation or
run-length encoding; LZ77 can represent prefixes which extend past the
end of the sliding window by ``feeding'' output characters back into
the decoder, thus allowing repetative cyclic prefixes to be
represented compactly.

% To illustrate: suppose we are to encode the tongue-twister ``can you
% can a can as a can man can can a can'' using LZ77, with a window of
% length 8. The string is 44 characters long, thus requiring 44 bytes to
% store. In its implementation, the algorithm implicitly holds a
% windows-worth full of null-character bytes (represented in text as
% \texttt{\textbackslash 0 or 0x00}) sets the current character to the
% first character of the input. Since, in this case, the first 7
% characters of the input are all unique, and as such they cannot be
% represented by duplicate prefixes from earlier in the text (as there
% is no ``earlier text'' which matches it.) Thus each of these first
% bytes is emitted as \texttt{(0, 1, character)}. By this point, the
% current prefix has advanced to the second space character. Since there
% is a space less than 8 characters before this, the next character
% (\texttt{'c'}) is appended to the current prefix. Since the prefix
% \texttt{' c'} is not found in the window, the algorithm emits
% \texttt{(4, 1, 'c')}. The next prefix is set to \texttt{'a'} and the
% process repeats. This time, the prefix \texttt{'an '} can be found in
% the window, so the algorithm proceeds up to the following \texttt{'c'}
% and then emits \texttt{(7, 3, 'c')}. The next prefix is set to
% \texttt{'a'} and the algorithm continues. For brevity, the remaining
% steps will not be listed, but the compressed version contains 17
% 3-tuples, requiring 53 bytes to store. This example shows not only
% that any compression algorithm has incompressable inputs, but also
% that the design of an algorithm can affect which inputs are
% incompressable. (In this case, the first few bytes are encoded as
% three-byte triples, whereas later bytes are encoded as pure
% prefixes. Since this example is short, the cost incurred by the first
% bytes is too high to be overcome by the later savings.)

@@

Since knowing the structure of a given input can allow the
construction of a more efficient data compression scheme, it is
worthwhile to study some compression scheme which is specifically
designed to store a particular type of data. In this section I will
describe the structure and function of the Free Lossless Audio Codec,
known more commonly by its acronym FLAC, a popular format for
high-fidelity audio storage. \cite{flac}

### 2018-12-10 presentation/presentation.tex
Update the paper and presentation.

\begin{frame}
  \subsection{Impossibility of perfect compression}
  A compression scheme is considered ``perfect'' if and only if its
  image is the set of all strings of length strictly less than that of
  the domain.
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

@@

\begin{frame}
  \section{Lempel-Ziv Based Algorithms}
  \subsection{LZ}
  \frametitle{Lempel-Ziv compression algorithms}
  LZ77 and LZ78 (a.k.a. LZ1 and LZ2, a.k.a. LZ)
  \begin{itemize}
  \item Introduced in 1977 and 1978 by Abraham Lempel and Jakob Ziv
  \item Dictionary block compressor family
  \item Uses a sliding-window approach
  \item Produces 3-tuples representing prefixes
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZ77 Example}
  Suppose we are to encode the tongue-twister ``Can you can a can as a
  canner can can a can?'' with a window size of 8.
\end{frame}

\begin{frame}
  \section{FLAC}
  \frametitle{Free Lossless Audio Codec}
  The Free Lossless Audio Codec is a domain-specific compression
  scheme for hi-fi audio.
\end{frame}

\end{document}

>>>>

\begin{frame}
  \subsection{Impossibility of Perfect Compression}
  A compression scheme is considered ``perfect'' if and only if its
  image is the set of all strings of length strictly less than that of
  the domain.
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

@@

\begin{frame}
  \section{Lempel-Ziv Based Algorithms}
  \subsection{LZ77 and LZ78}
  \frametitle{Lempel-Ziv compression algorithms}
  LZ77 and LZ78 (a.k.a. LZ1 and LZ2, a.k.a. LZ)
  \begin{itemize}
  \item Introduced in 1977 and 1978 by Abraham Lempel and Jakob Ziv
  \item Dictionary block compressor family
  \item Uses a sliding-window approach
  \item Produces 3-tuples representing prefixes
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZ77 Example}
  \frametitle{A Brief Example}

  Suppose we are to encode the word ``hahahahaha'' with a window size
  of 16.
\end{frame}

\begin{frame}
  \frametitle{LZ78}
  LZ78 is a slightly different modification on the LZ77 algorithm.
  \begin{itemize}
  \item Instead of a sliding window, it uses a ``chained dictionary''
  \item Each entry in the dictionary is a two-tuple: \texttt{(
      previous prefix index, following character)}
  \item The dictionary can be represented as a tree data structure
  \item Equivalent to LZ77 even though it uses a different method
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZW and DEFLATE}
  \frametitle{Derivative Methods: LZW and DEFLATE}

  Since they are very effective and efficient algorithms, LZ77 and
  LZ78 have inspired a wide variety of derivative methods.

  \begin{itemize}
  \item LZW (Lempel-Ziv-Welch): A method based on LZ78 which begins
    with a pre-populated prefix dictionary, with each individual
    character in the alphabet.
  \item DEFLATE: Splits the input into blocks, then encodes the
    alphabet using a Huffman code, then compresses the resultant input
    channel using LZ78.
  \end{itemize}
\end{frame}

\begin{frame}
  \section{FLAC}
  \frametitle{Free Lossless Audio Codec}
  The Free Lossless Audio Codec is a domain-specific compression
  scheme for hi-fi audio.
  \begin{enumerate}
  \item Split input into blocks (can optimize for compression or
    speed)
  \item Encode stereo channels (left-right or mid-side)
  \item Fit the sound signal to an n-degree polynomial (configurable)
  \item Calculate the residuals (difference between the raw signal and
    the fit polynomial)
  \item Encode the residual values using Rice codes (essentially a
    Huffman coding optimized for alphabets with geometric
  \end{enumerate}
\end{frame}

\begin{frame}
  \section{Question and Answer}
  \frametitle{Any questions?}
\end{frame}

\begin{frame}
  \frametitle{Thank you!}
\end{frame}
\end{document}
